---
execute: 
  warning: false
  eval: false
---

# Appendix: Data acquisition {.unnumbered}

Here, I will describe how I obtained each example data set used in this course.

```{r}
set.seed(123)

# Required packages
library(here)
library(tidyverse)
library(igraph)
```

## got.rds

This file contains an `igraph` object with a network consisting of character
relationships in George R. R. Martin's *A Storm of Swords*, the third novel
in his series *A Song of Ice and Fire*. Data were originally compiled by
@beveridge2016network.

```{r}
# Get edges
got_edges <- read_csv(
    "https://raw.githubusercontent.com/melaniewalsh/sample-social-network-datasets/master/sample-datasets/game-of-thrones/got-edges.csv",
    show_col_types = FALSE
) |>
    dplyr::rename(from = Source, to = Target, weight = Weight) |>
    as.data.frame()

# Get nodes
got_nodes <- read_csv(
    "https://raw.githubusercontent.com/melaniewalsh/sample-social-network-datasets/master/sample-datasets/game-of-thrones/got-nodes.csv",
    show_col_types = FALSE
) |>
    dplyr::rename(node = Id, label = Label) |>
    as.data.frame()

# Create igraph object
got <- graph_from_data_frame(
    got_edges, directed = FALSE, vertices = got_nodes
)

saveRDS(
    got, compress = "xz", file = here("data", "got.rds")
)
```


## References {.unnumbered}

